---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# butterflyGamSims


## Installation

You can install the development version of butterflyGamSims from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cbedwards/butterflyGamSims")
```

## Description

**butterflyGamSims** can be used to simulate butterfly transect data with known properties,
fit possible generalized additive models to estimate activity curves, calculate abundance index and
phenology metrics from these estimated activity curves, and using one or more data filtering
protocols, estimate across-year trends in abundance and phenology. My primary motivation for creating this package was to compare estimated and real trends in abundance and phenology for a range of gam approaches and data censoring options to identify potential biases when fitting real data. In particular, in exploring real data, I found that for sufficiently sparse data in which abundance declined across a time series, apparent trends in phenology became unreliable without 
some form of removing years with insufficient data. This package allows systematic exploration of
*what* rules for removing data are sufficient to avoid notable bias.

For this purpose, `butterfly_gam_sim()` and `butterfly_load()` are the key functions.
`butterfly_gam_sim()` takes all the arguments needed to define the data to be simulated,
the gam-fitting parameters, and the data filtering protocol. See `?butterfly_gam_sim` for details on all the arguments (and a working example). `butterfly_load()` reads in the files saved by a
run of `butterfly_gam_sim()` and organizes them into a single list; `?butterfly_load` contains
all the relevant information for interpretting the simulated time series, estimated yearly metrics,
and estimated trends.

Because **butterflyGamSims** necessarily simulates butterfly time series data, the 
timeseries simulation function `timeseries_sim()` may be helpful for other projects (ie
simulated data for teaching/training, other methods testing projects).
Note that `timeseries_sim()` returns a set of simulated time series; depending on your needs,`timeseries_gen()` (which saves them to a file) may be more helpful.

For visualizing simulated timeseries and associated fitted gam models, `timeseries_examples()` and `fit_plotter()` may be helpful. As a simple example, 
here we simulate a 10-year time series with a gaussian underlying activity curve centered
on day 130 with standard deviation 15, a starting abundance index of 500, a moderate
exponential decline in population across years, and 7 census per year with poisson sampling error.

```{r}
library(butterflyGamSims)
timeseries_examples(years = 1990:2000,
                    doy.samples = seq(105,160, by = 7),
                    abund.type = "exp",
                    activity.type = "gauss",
                    sample.type = "pois",
                    growth.rate = -0.12,
                    init.size = 500,
                    act.mean = 130,
                    act.sd = 15)
```


And here we fit a similar timeseries with a simple gam (using a tensor product smooth with a cubic regression smooth in both the day-of-year and the year dimensions),

```{r}
dat.sim = timeseries_generator(years = 1990:2000,
                               doy.samples = seq(105,160, by = 7),
                               abund.type = "exp",
                               activity.type = "gauss",
                               sample.type = "pois",
                               growth.rate = -0.12,
                               init.size = 500,
                               act.mean = 130,
                               act.sd = 15)
out = gam_fitter(years.vec = dat.sim$years,
                 doy.vec = dat.sim$doy,
                 count.vec = dat.sim$count,
                 doy.smooth = "cr",
                 doy.knots = 5,
                 years.smooth = "cr",
                 anchor.flag = FALSE
)
fit_plotter(dat.timeseries = dat.sim,
            dat.fitted = out$dat.fitted,
            activity.curve = out$activity.curve)
```


