---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# butterflyGamSims


## Installation

You can install butterflyGamSims from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cbedwards/butterflyGamSims")
```

## Description

### Motivation

In ecology and in conservation biology, understanding the population trends and timing of activity ("phenology") for insects is an ongoing interest and/or concern. Particularly in the context of climate change, ecologists are interested in identifying trends in phenology (ie populations shifting to become
active earlier, or later, or changing length of time they are active within the year). These shifts in phenology are one of the clearest signals of climate change on populations ([Parmesan 2006](https://www.jstor.org/stable/30033846)), and there concerns that shifts in phenology could lead to mismatches between populations and their food sources ([Miller-Rushing et al. 2010](https://doi.org/10.1098/rstb.2010.0148)), or lead to "developmental traps" in which populations try to squeeze in an additional generation that fails to develop before conditions become too cold ([Kerr et al. 2018]( https://doi.org/10.1111/gcb.14959)).

Because of their visibility and public interest, there is a long history of gathering data on butterflies (for example, the [UKBMS](https://ukbms.org/) is a systematic butterfly monitoring program that has been running since 1976). The richness of butterfly data allows us both to explore the specifics of individual butterfly species and to use them as a general proxy for possible trends in insects in general (although butterflies are an imperfect stand-in for other insects, and there is a ton of great work going into systematic surveys of other taxa!). Given the relative wealth of butterfly data, it's important to have tools to translate butterfly observations into estimates of abundance and phenology.

One of the most common forms of systematic butterfly data are repeated transect counts, where scientists or volunteers walk a fixed route multiple times per year and report the number of butterflies they saw of one or more species on each day that they censused. (The specific methods can vary -- initial protocols were developed by Pollard ([Pollard 1977](https://doi.org/10.1016/0006-3207(77)90065-9)), but there have been elaborations since.). As an example of the depth of data available, check out [Pollardbase](https://www.pollardbase.org/), in which [Prof. Leslie Ries](https://www.butterflyinformatics.org/person/leslie-ries) have taken on the Herculean task of aggregating, cleaning, and integrating repeated transect counts from programs across the US. There are a number of tools for translating these repeated transect counts to estimates of abundance or phenology, including Bayesian approaches like [Lindén and Mäntyniemi 2011](https://doi.org/10.1890/10-1831.1) or [Champan et al. 2015](https://doi.org/10.1111/bij.12515), fitting Gaussian curves with regression models like [Edwards and Crone 2021](https://doi.org/10.1111/oik.08368), or fitting the Zonneveld model [Zonneveld 1991]( https://doi.org/10.1111/j.1365-2311.1991.tb00198.x) perhaps with the [INCA](https://www.urbanwildlands.org/INCA/) tool. One of the most flexible approaches is to fit smoothing splines (a form of generalized additive models (GAMs)) to the data in order to estimate the yearly activity curves, from which we can estimate abundance index and phenology metrics (e.g. [Hodgson et al. 2011](https://doi.org/10.1111/j.1365-2486.2010.02308.x), [Wepprich et al. 2019](https://doi.org/10.1371/journal.pone.0216270), [Stemkovski et al. 2020](https://doi.org/10.1111/ele.13583)). The flexibility of smoothing splines is particularly valuable when working with species that have more than one generation per year, or might have unusual activity curves (for example, when male and female become active at different times). However, that flexibility comes at a cost, and when data is sparse, GAMs can predict unreasonable acitivy periods that lead to incorrect and/or biased estimates of abundance or phenology. At the same time, it's vital that we understand the dynamics of rare species and small populations: among other things, only working with common species would bias our understanding of overall trends across species ([Didham et al. 2020]( https://doi.org/10.1111/icad.12408)). 

**Long story short**, we need to be able to test how reliable different variations of GAM models are, and if there are rules for excluding years with insufficient data that balance our need for reliable estimates of abundance and phenology with our desire not to exclude years or species with less data. This package simulates transect data for biologically reasonable butterfly activity with reasonable reasonable sampling variation, then fits the data with the specified gam model and estimates trends in abundance and phenology across years, using the specified rules for excluding potential problem years. Because the data is simulated, we know what the *true* trends in the population are, so we can evaluate how well our methods worked. 



### Using this package

My primary motivation for creating this package was to compare estimated and real trends in abundance and phenology for a range of gam approaches and data censoring options to identify potential biases when fitting real data. In particular, in exploring real data, I found that for sufficiently sparse data in which abundance declined across a time series, apparent trends in phenology became unreliable without 
some form of removing years with insufficient data. This package allows systematic exploration of
*what* rules for removing data are sufficient to avoid notable bias.

For this purpose, `butterfly_gam_sim()` and `butterfly_load()` are the key functions.
`butterfly_gam_sim()` takes all the arguments needed to define the data to be simulated,
the gam-fitting parameters, and the data filtering protocol. See `?butterfly_gam_sim` for details on all the arguments (and a simple example). `butterfly_load()` reads in the files saved by a
run of `butterfly_gam_sim()` and organizes them into a single list; `?butterfly_load` contains
all the relevant information for interpreting the simulated time series, estimated yearly metrics,
and estimated trends.

Because **butterflyGamSims** necessarily simulates butterfly time series data, the 
timeseries simulation function `timeseries_sim()` may be helpful for other projects (ie
simulated data for teaching/training, other methods testing projects).
Note that `timeseries_sim()` returns a set of simulated time series; depending on your needs,`timeseries_gen()` (which saves them to a file) may be more helpful.

For visualizing simulated timeseries and associated fitted gam models, `timeseries_examples()` and `fit_plotter()` may be helpful. As a simple example, 
here we simulate a 10-year time series with a gaussian underlying activity curve centered
on day 130 with standard deviation 15, a starting abundance index of 500, a moderate
exponential decline in population across years, and 7 census per year with poisson sampling error.

```{r}
library(butterflyGamSims)
timeseries_examples(years = 1990:2000,
                    doy.samples = seq(105,160, by = 7),
                    abund.type = "exp",
                    activity.type = "gauss",
                    sample.type = "pois",
                    growth.rate = -0.12,
                    init.size = 500,
                    act.mean = 130,
                    act.sd = 15)
```


And here we fit a similar timeseries with a simple gam (using a tensor product smooth with a cubic regression smooth in both the day-of-year and the year dimensions),

```{r}
dat.sim = timeseries_generator(years = 1990:2000,
                               doy.samples = seq(105,160, by = 7),
                               abund.type = "exp",
                               activity.type = "gauss",
                               sample.type = "pois",
                               growth.rate = -0.12,
                               init.size = 500,
                               act.mean = 130,
                               act.sd = 15)
out = gam_fitter(years.vec = dat.sim$years,
                 doy.vec = dat.sim$doy,
                 count.vec = dat.sim$count,
                 doy.smooth = "cr",
                 doy.knots = 5,
                 years.smooth = "cr",
                 anchor.flag = FALSE
)
fit_plotter(dat.timeseries = dat.sim,
            dat.fitted = out$dat.fitted,
            activity.curve = out$activity.curve)
```

### A quick note on smoothing splines and transect data

For those who are not familiar, generalized additive models are powerful, but their implementation and interpretation can be a bit tricky. I use the fantastic R package [**mgcv**](https://cran.r-project.org/web/packages/mgcv/index.html) by Simon Wood ([Wood 2017](https://doi.org/10.1201/9781315370279)), and highly recommend [Noam Ross](https://github.com/noamross)'s [free online course in fitting smoothing splines with MGCV](https://noamross.github.io/gams-in-r-course/).

As a starting point for working with potentially sparse transect data, I recommend using a tensor product smooth across day of year ("doy") and year, which might be specified with the term `te(doy, year)`. The tensor product smooth across doy and year is representing our expectation that
- activity curves within a year will be smooth (smoothing across doy)
- sequential years should have relatively similar abundances (smoothing across year)
- sequential years  should have relatively similar phenologies and shapes of activity curves (the interaction part of `te()`)

The default for a 2-dimensional smooth with the `s(doy, year)` nomenclature assumes that wiggliness will be the same in both dimesions (here "doy" and "year" are our dimensions). This assumption generally only makes sense if our units are the same in both dimension, which is definitely not the case here. 

There are certainly other ways to represent variation across doy and years, but I have found them to be less less appropriate in this context. With sufficient data, fitting each year separately (either with separate gam model calls, or using the `by = year` argument in the smooth) makes each yearly activity curve independent, which can better capture the highly variable activity of insects. However, for sparse data (which is what I'm trying to tackle here), this prevents data-rich years from informing data-poor years, and leads to many more unreasoanble gam fits. When thinking about data-rich years informing data-poor years, it may be tempting to include year as a random effect in some fashion. However, this would lead to data-poor years being biased towards the average of more data-rich years (due to the shrinkage of random effects). This will tend to reduce apparent trends across year (since some years will be shrunk towards the average year), which is a big problem when our goal is to estimate trends. 

**butterflyGamSims** is designed to use GAMs fit with mgcv using the tensor product smooth across doy and year described above. The package allows users to specific some variations within this framework, including different types of smooth in each dimension (cubic regression smooths are a common default, but cyclic smooth in the doy direction can enforce our expectation that Dec 31 and Jan 1 should be similar), the number of knots in the doy direction (more knots = more flexibility, which is better for capturing real variation in the data but risks overfitting), and the addition of "anchor" zeroes (adding in additional 0-observation entries some distance before and after all real data points to represent our understanding that biological activity is unreasonable in certain seasons).
